---
title: 'Lab 2 : Data Visualization with ggplot'
---

## Learning objectives

-   Instailling R packages
-   Built-in R data sets and data set packages
-   ggplot2

## Introduction to R Graphics

R provides comprehensive graphics utilities for visualizing and exploring scientific data. To date we have been making a few plots using the R Base Graphics. In addition, several more recent graphics environments extend these utilities. These include the `grid`, `lattice` and `ggplot2` packages. All have the roles, but `ggplot2` environment that is part of the Tidyverse package has become popular and is now used for many R packages and in scientific publications.

### ggplot2 and the Grammar of Graphics

`ggplot2` is meant to be an implementation of the Grammar of Graphics, hence the gg in ggplot. The basic notion is that there is a grammar to the composition of graphical components in statistical graphics. By directly controlling that grammar, you can generate a large set of carefully constructed graphics from a relatively small set of operations. As Hadley Wickham (2010), the author of `ggplot2` said,

"A good grammar will allow us to gain insight into the composition of complicated graphics, and reveal unexpected connections between seemingly different graphics."

That said learning good gradatmmar can be challenging, especially when the computer is catching your mistakes.

### 

https://ggplot2.tidyverse.org/index.html

### Tutorials and resources

You can make amazing graphs with ggplot, but there is a long learning curve so we will have multiple lab sessions on ggplot and graphing. Here are a few different resources for `ggplot`.

-   Hadley Wickham and Garrett Grolemund released the second edition of [R for Data Science](https://r4ds.hadley.nz/).
-   Data Carpentry's [Data Analysis and Visualization in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/)
-   For those with a visual learning style there is [Maria Nattestad's Youtube videos](https://www.youtube.com/channel/UC2bWYX9h1KlaGWfTDuhASWg)
-   The [ggplot cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)

## On the Computer

### Create and save your Quarto Markdown (qmd) file

Just like last week we will be writing our code in a Quarto Markdown (qmd) file. Remember to use the following formatting in your YAML block

```         
---
title: "Lab 2 Data Visualization"
author: "You"
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
editor: visual
---
```

### Installing and loading R packages

In this course we will work with many different R packages that will need to fbe installed on your computer. I have already installed most of these packages for students on Posit Cloud. If you are working on your own computer or on Unity, you can install them using Tools > Install Packages. You only need to install a package once!

To work with an R package load it with the `library` command. I always load my packages at the beginning of my files.

```{r}
#| warning: false
# The above warning: false command is used to suppress
# known warning messages when rendering the qmd file

library(tidyverse)
```

### Data for today's lab

In most labs we will be loading in data from files (e.g. our 23andME SNP data). Today for simplicity we will work with data sets that come with R and the are available as R packages.

#### Data sets (data frames) that come with R

R contains pre-loaded data sets that will see in many examples posted on the internet. The `mtcars` and `iris` data sets are very popular. You can see the whole list by typing data(). This will pop up a window with a list of the data sets.

```{r}
#| eval: false
# The above eval: false statement is used because it will result
# in an error when rendering the qmd file. 

data()
```

In class we will talk more about the structure of a data set, which can be summarized using the `str` command

```{r}
glimpse(iris)
```

You can see the whole data set by typing the name `iris` or by typing `view(iris)` which will pop up a window with the data set. However we don't want to show all 150 observations (rows) of the iris data set in this document. We can use the `head` command to show just the first 5 rows.

```{r}
head(iris)
```

#### Data sets that are part of R packages

R for Data Science uses the `palmerpenguins` package, "which includes the penguins dataset containing body measurements for penguins on three islands in the Palmer Archipelago, and the ggthemes package, which offers a colorblind safe color palette. We will load these for our work today."

```{r}
library(palmerpenguins)
library(ggthemes)
```

Data Analysis and Visualization in R for Ecologists uses the `ratdat` package, a long-term dataset from Portal, Arizona, in the Chihuahuan desert.

```{r}
library(ratdat)
```

The help command can be used to learn more about the `palmerpenguins` and `ratdat` packages. After running the below commands, in the right bottom corner under the `Help` tab the package documentation can be viewed

```{r}
#| eval: false
help(package="palmerpenguins")
help(package="ratdat")
```

### [R for Data Science Chapter 1](https://r4ds.hadley.nz/data-visualize). 

Today we will walk through Chapter 1 of R for Data Science. By putting the examples and exercises in our own Quarto Markdown file, we can create own personal path through the Chapter. We will do all sections, except `1.6 Saving your plots` which we will learn later. In section `1.3` you only need to do exercises 1-5, 9 and 10. Make are readable report by delineating the sections (e.g. 1.2.3 Creating a ggplot) with hashtags so they are visible in your report outline.

## What to upload to Canvas

After you `Render` the qmd file to an html file, export the file to your computer and upload it to Canvas.


